Analyzing DECREASE trials for extent of data fabrication

### 2017-02-15 submission/manuscript.tex
Re-estimate inversion method with seed, revising through step 2

\setlength{\droptitle}{-2em}
  \title{Analyzing DECREASE trials for extent of data fabrication}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{Chris HJ Hartgerink, Gerben ter Riet, Marleen Kemper}
  \preauthor{\centering\large\emph}
  \postauthor{\par}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{01 February, 2017}

@@

The effect of beta-blockers on perioperative mortality in non-cardiac
surgery for patients with coronary artery disease (CAD) has been subject
of discussion\textsuperscript{1} due to findings of research
misconduct\textsuperscript{2--4} that were central to claims of their
effectiveness\textsuperscript{5--7}. Perioperative mortality is the
deathrate of patients during the period of the surgical procedure, which
typically includes admission, anaesthesia, surgery, and recovery. This
related finding of research misconduct altered meta-analytic results of
perioperative mortality in non-cardiac surgery for CAD patients
substantively\textsuperscript{7}, where meta-analyses that include these
studies conclude that beta-blockers decrease perioperative
mortality\textsuperscript{5--7} but a meta-analysis that excludes these
studies concludes that beta-blockers increase perioperative
mortality\textsuperscript{7}.

The trials deemed to be the result of research misconduct were the Dutch
DECREASE trials\textsuperscript{2--4}. The committees that investigated
the integrity of the DECREASE studies reported that data fabrication was
likely\textsuperscript{2--4}, but that the extent of the data
fabrication remained unclear. Given that the inclusion or exclusion of
these trials in meta-analyses is crucial in determining whether
beta-blockers are indeed effective during the perioperative period and
the latest guidelines still recommend the usage of
beta-blockers\textsuperscript{8}, we aim to estimate the extent of data
fabrication in the DECREASE studies\textsuperscript{9,10} in order to
further this debate.

The reports on the DECREASE trials primarily focused on the provenance
of the raw data and patient files (e.g., informed consent, whether data
corresponded to patient files), but neglected the extent to which the
DECREASE studies deviated from similar trials that tested the
effectiveness of beta-blockers on perioperative mortality. Comparing how
(dis)similar trials are from comparable studies is not a new
idea\textsuperscript{11} and has proven effective in detecting data
fabrication\textsuperscript{12}. This is a different approach from the
forensic statistical methods previously applied\textsuperscript{4},
where results within one study are evaluated as opposed to other similar
studies. This explains the discrepancy between the feasibility of
evaluating the DECREASE trials by the statistical expert of the
committee report and ours.

To statistically investigate the evidence of data fabrication in the
results of the DECREASE studies\textsuperscript{9,10}, we took three
steps. First, we replicate the findings from a 2014
meta-analysis\textsuperscript{7} that contains sufficient information to
estimate the deviation of the DECREASE-I and DECREASE-IV studies from
the remaining studies. Second, we evaluate the veracity of the
DECREASE-I and DECREASE-IV studies (i.e., the probability that these
study findings occurred assuming no data fabrication occurred). Third,
we reverse this assumption and assume that data fabrication did occur
and estimate how many data points would have to be fabricated to
reproduce the results of the DECREASE-I and DECREASE-IV studies if the
remaining studies are regarded as estimating the true effect of
beta-blockers on perioperative mortality.

@@

To ensure that we used similar analysis procedures as in the 2014
meta-analysis\textsuperscript{7}, we initially reproduced Bouri et al.'s
estimates. This ensured that (1) their results are reproducible and (2)
we are using the correct estimates in subsequent steps of our analyses.
Using figures 2 and 3 from the original paper\textsuperscript{7}, we
extracted the raw event data for the 2 (control vs experimental) by 2
(event or no event) design, which we used to recompute the natural
logarithm of the risk ratio and its standard error. The extracted event
data is available at \href{https://osf.io/aykeh}{osf.io/aykeh} and our
analysis plan was preregistered at
\href{https://osf.io/vnmzc}{osf.io/vnmzc}.

@@

Second, we meta-analyzed all studies combined, including a dummy
predictor for the DECREASE and non-DECREASE studies to replicate results
presented in Figure 4 of the 2014 meta-analysis\textsuperscript{7}.
Surprisingly, our results showed more evidence for unequal subgroups
than the original meta-analysis\textsuperscript{7}; whereas the original
test for subgroup differences was \(\chi^2(1)=3.91,p=.05\), our analyses
indicated that \(\chi^2(1)=6.12,p=0.013\). Additionally, the original
analyses showed substantial heterogeneity (\(I^2=74.4\)\%) where we
found an \(I^2=0\)\%. Different heterogeneity estimators (e.g.,
\texttt{DerSimonian-Laird} instead of \texttt{REML}) did not resolve
this difference. We tried to clarify these discrepancies by e-mailing
the original authors, but failed to receive a response.

@@

We estimated the number of data points that would need to be fabricated
to arrive at the estimates from the DECREASE studies, given that the
non-DECREASE studies represent the true effect. This assumes, in
contrast to Step 2, that the DECREASE studies in fact contain fabricated
data. The estimates from Step 3 provide an indication of the extent of
the potential data fabrication in the DECREASE studies under this
assumption\textsuperscript{2--4,7}.

@@

In order to estimate the number of fabricated data points, we estimated
the population effect distribution of mortality (in log odds) per
condition in both the DECREASE- and non-DECREASE studies separately. For
each of these, we ran a meta-analysis applying the same, verified
methods used in Step 1. These separate analyses were necessary,
considering that Step 1 only provided estimates of the overall risk
ratio but not the odds conditional on exposure (beta-blocker or control)
and study set (DECREASE- versus non-DECREASE). This resulted in four
meta-analytic mortality estimates with an accompanying variance.
Throughout the simulations, we use the point estimates to simulate
genuine- and fabricated data, but supplement this by using the
distribution estimates as sensitivity analyses.

We applied the inversion method to estimate the number of fabricated
data points in the DECREASE-I and DECREASE-IV
studies\textsuperscript{17}. With the inversion method, we iteratively
hypothesized that \emph{X} out of \emph{N} data points were fabricated
(i.e., \(0, 1, ..., N\)) in the DECREASE-I and DECREASE-IV studies
separately. For each study, we simulated 10,000 datasets per hypothesis
of \emph{X} fabricated data points and \emph{N-X} genuine data points.
For each simulated dataset (exact simulation procedure in next
paragraph), we determined the likelihood of the results with

@@

where \(\pi_{E}\) indicates the mortality rate of the experimental
condition as drawn from the meta-analytic effect distribution
(\(\pi_{C}\) indicates the same for the control condition). The
likelihood was computed under both the fabricated effect estimates
(i.e., \(L_{fabricated}\)) and the genuine data (i.e., \(L_{genuine}\)).
Table 2 indicates which cell sizes the various \(n_{XX}\) refer to
within the (simulated) data. We compared the likelihoods to determine
whether the simulated data were more likely to arise from the genuine-
(\(L_{genuine}\)) or from the fabricated data (\(L_{fabricated}\)). Note
that using the likelihoods for this is a minor deviation from the
preregistration, where we initially planned on using \(p\)-value
comparisons (\href{https://osf.io/vnmzc}{osf.io/vnmzc}).

@@

For each hypothesis of \emph{X} out of \emph{N} fabricated datapoints,
we computed the probability that the fabricated data are more likely
than the genuine data (\(p_F=P(L_{fabricated}>L_{genuine})\)). Based on
\(p_F\), we computed the confidence interval for \(X\) (i.e.,
\(X_{LB};X_{UB}\)). For a 95\% confidence interval, the lowerbound is
equal to the \(p_F\) closest to .025, whereas the upperbound is equal to
the \(p_F\) closest to .975.

We computed \(p_F\) for all possibilities of \emph{X} out of \emph{N}
fabricated datapoints in 10,000 randomly generated datsets, which were
generated in three steps. For each dataset we:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Sampled (without replacement) \emph{X} fictitious participants that
  would be the result of data fabrication (sampled across all
  conditions).
\item
  Determined the population mortality rate for each condition (i.e., for
  each cell as in Table 3). The meta-analytic point estimate was used or
  a population effect was randomly drawn from the meta-analytic effect
  distribution.
\item
  Simulated the number of deaths for the different conditions using a
  binomial distribution based on the mortality rate as determined in 2,
  resulting in the cell counts as in Table 3.
\end{enumerate}

Based on the meta-analytic effect from 2 and the cell sizes from 3, we
computed the likelihoods of \(L_{fabricated}\) and \(L_{genuine}\) using
Equation \ref{eq1}. As mentioned before, \(p_F\) was computed as the
probability that the data were more likely under the estimates resulting
from the (allegedly) fabricated data (i.e., the DECREASE trials) than
under the estimates resulting from the genuine data (i.e., the
non-DECREASE trials; \(p_F=P(L_{fabricated}>L_{genuine})\)).

@@

For DECREASE-I, the 95\% confidence interval for the estimated number of
fabricated data points is {[}0 - 112{]} or {[}3 - 112{]} when based on a
point estimate or a more uncertain distribution estimate, respectively.
The left column of Figure 3 depicts the \(p_F\) per \emph{X} fabricated
data points (top panel) and the bounds of the confidence interval when
the degree of confidence is altered (lower panel). Staying clearly
between the dotted lines in the top panel, depicting the 95\% CI (top:
.975; bottom: .025), it becomes apparent that the degree of uncertainty
is too high to make any estimates with a high degree of confidence. This
due to the small sample size (i.e., \(N=112\)) combined with the
availability of just the summary results. Only when the degree of
confidence is lowered to around 75\% does the interval not span the
entire sample size. As such, based on the summary results, little can be
said about the extent of the data fabrication that occurred in the
DECREASE-I trial, affirming the conclusions of the original committee
report\textsuperscript{4}.

For DECREASE-IV, the 95\% confidence interval for the estimated number
of fabricated data points is {[}8 - 1059{]} or {[}7 - 1066{]} when based
on a point estimate or a more uncertain distribution estimate,
respectively. The minor difference between the estimates indicates that
there is a high degree of confidence that data fabrication did occur.
Nonetheless, the range of potentially fabricated data points is still
estimated at approximately 1000; this indicates that the summary results
are insufficient to provide more than an estimated lowerbound. However,
this does indicate that it is possible not all data were fabricated
(i.e., \(N=1066\)), increasing the importance of data provenance to
discern between genuine and falsified data. This affirms the conclusion
of the scientific integrity committee that the results of the
DECREASE-IV trial are ``scientifically incorrect''
(p.11)\textsuperscript{3}.

The results also highlight that, despite the lack of availability of raw
data, summary results from larger samples result in more certainty about
the estimated number of fabricated data points using the inversion
method. This is due to decreased standard errors at the population level
of the effects, providing less variation in the simulations and
resulting in higher sensitivity. However, this also highlights that in
order to prevent detection it would be in the fabricators' interest to
fabricate small studies (assuming the fabricator wants to remain
undetected).

>>>>

\setlength{\droptitle}{-2em}
  \title{Analyzing DECREASE trials for extent of data fabrication}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{Chris HJ Hartgerink, Gerben ter Riet, Marleen Kemper}
  \preauthor{\centering\large\emph}
  \postauthor{\par}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{13 februari, 2017}

@@

The effect of beta-blockers on perioperative mortality in non-cardiac
surgery for patients with coronary artery disease (CAD) has been subject
of discussion\textsuperscript{1} due to findings of research
misconduct\textsuperscript{2--4}. Consequently, it appeared that the
related trials were central to claims of the effectiveness of
beta-blocker in decreasing perioperative
mortality\textsuperscript{5--7}. Perioperative mortality is the
deathrate of patients during the period of the surgical procedure, which
typically includes admission, anaesthesia, surgery, and recovery.
Meta-analyses that include these studies subject to research misconduct
conclude that beta-blockers decrease perioperative
mortality\textsuperscript{5--7} whereas a meta-analysis that excludes
the studies subject to research misconduct concludes that beta-blockers
increase perioperative mortality\textsuperscript{7}.

The trials subject to research misconduct were the Dutch DECREASE
trials\textsuperscript{2--4}. The committees that investigated the
integrity of the DECREASE studies reported that data fabrication was
likely but that the extent of the data fabrication remained
unclear\textsuperscript{2--4}. Given that the inclusion or exclusion of
these trials in meta-analyses is crucial in determining whether
beta-blockers are an effective treatment and considering that the latest
guidelines still recommend the usage of
beta-blockers\textsuperscript{8}, we aim to estimate the extent of data
fabrication in the DECREASE studies\textsuperscript{9,10} in order to
further the debate.

The reports on the integrity of the DECREASE trials primarily focused on
the provenance of the raw data and patient files (e.g., informed
consent, whether data corresponded to patient files) but did not
investigate the extent to which the DECREASE trials deviated from
comparable trials. The committee reports did not neglect statistical
evaluation; a statistical expert evaluated the applicability of forensic
statistical methods\textsuperscript{4} to evaluate results within one
trial. However, comparing across trials can be used as a forensic
statistical method as well. This method has previously been used to
monitor trial data quality or to test for potential data
anomalies\textsuperscript{11}. Moreover, this method has previously
proven to be effective in detecting data
fabrication\textsuperscript{12}. As such, evaluating the DECREASE trials
against comparable trials investigating the effectiveness of
beta-blockers with respect to perioperative mortality could prove
worthwhile.

To statistically investigate the evidence of data fabrication in the
results of the DECREASE studies\textsuperscript{9,10}, we took three
steps. First, we reproduced the findings from a 2014
meta-analysis\textsuperscript{7} that contained sufficient information
to estimate the deviation of the DECREASE-I and DECREASE-IV studies from
the remaining studies. Second, we evaluated the veracity of the
DECREASE-I and DECREASE-IV studies (i.e., the probability that these
study findings occurred assuming no data fabrication occurred). Third,
we reversed this assumption and assumed that data fabrication did occur
and estimate how many data points would have to be fabricated to
reproduce the results of the DECREASE-I and DECREASE-IV studies if the
remaining studies are regarded as estimating the true effect of
beta-blockers on perioperative mortality.

@@

To ensure that we used similar analysis procedures as in the 2014
meta-analysis\textsuperscript{7}, we initially reproduced Bouri et al.'s
estimates. This ensured that (1) their results are reproducible and (2)
we are using the correct estimates in subsequent steps of our analyses.
Using figures 2 and 3 from the original paper\textsuperscript{7}, we
extracted the raw event data for the 2 (control vs experimental) by 2
(event vs no event) design, which we used to recompute the natural
logarithm of the risk ratio and its standard error. The extracted event
data is available at \href{https://osf.io/aykeh}{osf.io/aykeh} and our
analysis plan was preregistered at
\href{https://osf.io/vnmzc}{osf.io/vnmzc}.

\begin{verbatim}
## Warning: package 'latex2exp' was built under R version 3.2.5
\end{verbatim}

\begin{verbatim}
## Warning: package 'knitr' was built under R version 3.2.5
\end{verbatim}

@@

Second, we meta-analyzed all studies combined, including a dummy
predictor for the DECREASE and non-DECREASE studies to reproduce results
presented in Figure 4 of the 2014 meta-analysis\textsuperscript{7}.
Surprisingly, our results showed more evidence for unequal subgroups
than the original meta-analysis\textsuperscript{7}; whereas the original
test for subgroup differences was \(\chi^2(1)=3.91,p=.05\), our analyses
indicated that \(\chi^2(1)=6.12,p=0.013\). Additionally, the original
analyses showed substantial heterogeneity (\(I^2=74.4\)\%) where we
found an \(I^2=0\)\%. Different heterogeneity estimators (e.g.,
\texttt{DerSimonian-Laird} instead of \texttt{REML}) did not resolve
this difference. We tried to clarify these discrepancies by e-mailing
the original authors, but did not receive a response.

@@

We estimated the number of data points that would need to be fabricated
to arrive at the estimates from the DECREASE studies, given that the
non-DECREASE studies represent the true population effect distribution.
This assumes, in stark contrast to Step 2, that the DECREASE studies in
fact contain fabricated data. The estimates from Step 3 provide an
indication of the extent of potential data fabrication in the DECREASE
studies\textsuperscript{2--4,7}.

@@

In order to estimate the number of fabricated data points, we estimated
the population effect distribution of mortality (in log odds) per
condition in both the DECREASE- and non-DECREASE studies. For each of
these, we ran a meta-analysis applying the same, verified methods used
in Step 1. These separate analyses were necessary, considering that Step
1 only provided estimates of the overall risk ratio but not the odds
conditional on exposure (beta-blocker or control) and trial set
(DECREASE- versus non-DECREASE). This resulted in four meta-analytic
mortality estimates with an accompanying effect variance. Throughout the
simulations, we use the point estimates to simulate genuine- and
fabricated data but supplement this by using the more uncertain
distribution estimates as sensitivity analyses.

We applied the inversion method to estimate the number of fabricated
data points in the DECREASE-I and DECREASE-IV
trials\textsuperscript{17}. With the inversion method, we iteratively
hypothesized that \emph{X} out of \emph{N} data points were fabricated
(i.e., \(X={0, 1, ..., N}\)). For each combination of \emph{X} and
trial, we simulated 10,000 datasets. Each simulated dataset contained
\emph{X} fabricated data points and \emph{N-X} genuine data points. For
each simulated dataset (exact simulation procedure in next paragraph),
we determined the likelihood of the results with

@@

where \(\pi_{E}\) indicates the mortality rate of the experimental
condition as drawn from the meta-analytic effect distribution
(\(\pi_{C}\) indicates the same for the control condition). The
likelihood was computed under both the fabricated effect estimates
(i.e., \(L_{fabricated}\)) and the genuine data (i.e., \(L_{genuine}\)).
Table 2 indicates which cell sizes the various \(n_{XX}\) refer to
within the (simulated) data. Subsequently, we compared the likelihoods
to determine whether the simulated data were more likely to arise from
the genuine trials (\(L_{genuine}\)) or from the fabricated trials
(\(L_{fabricated}\)). Note that using the likelihoods for this is a
minor deviation from the preregistration, where we initially planned on
using \(p\)-value comparisons
(\href{https://osf.io/vnmzc}{osf.io/vnmzc}).

@@

For each hypothesis of \emph{X} out of \emph{N} fabricated data points,
we computed the probability that the fabricated data are more likely
than the genuine data (\(p_F=P(L_{fabricated}>L_{genuine})\)). Based on
\(p_F\), we computed the confidence interval for \(X\) (i.e.,
\(X_{LB};X_{UB}\)). For a 95\% confidence interval, the lowerbound is
equal to the \(p_F\) closest to .025, whereas the upperbound is equal to
the \(p_F\) closest to .975.

We computed \(p_F\) for all \emph{X} out of \emph{N} fabricated
datapoints in 10,000 randomly generated datsets, which were generated in
three steps. For each dataset we:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Sampled (without replacement and across conditions) \emph{X}
  fictitious participants that would be the result of data fabrication.
\item
  Determined the population mortality rate for each condition (i.e., for
  each cell as in Table 3). The meta-analytic point estimate was used or
  a population effect was randomly drawn from the meta-analytic effect
  distribution.
\item
  Simulated the number of deaths for the different conditions using a
  binomial distribution based on the mortality rate as determined in 2,
  resulting in the cell counts as in Table 3.
\end{enumerate}

Based on the meta-analytic effect from 2 and the cell sizes from 3, we
computed the likelihoods of \(L_{fabricated}\) and \(L_{genuine}\) using
Equation \ref{eq1}. As mentioned before, we computed \(p_F\) as the
probability that the data were more likely under the estimates resulting
from the (allegedly) fabricated data (i.e., the DECREASE trials) than
under the estimates resulting from the genuine data (i.e., the
non-DECREASE trials; \(p_F=P(L_{fabricated}>L_{genuine})\)).

@@

For DECREASE-I, the 95\% confidence interval for the estimated number of
fabricated data points is {[}0 - 112{]} or {[}3 - 112{]} when based on a
point estimate or a more uncertain distribution estimate, respectively.
The left column of Figure 3 depicts the \(p_F\) per \emph{X} fabricated
data points (top panel) and the bounds of the confidence interval when
the degree of confidence is altered (lower panel). Staying clearly
between the dotted lines in the top panel, depicting the 95\% CI (top:
.975; bottom: .025), it becomes apparent that the degree of uncertainty
is too high to make any estimates about fabricated data points with a
high degree of confidence. This is partly due to the small sample size
(i.e., \(N=112\)) and the availability of just the summary results. Only
when the degree of confidence is lowered to around 75\% does the
interval not span the entire sample size. As such, based on the summary
results, little can be said about the extent of the data fabrication
that occurred in the DECREASE-I trial, affirming the conclusions of the
original committee report\textsuperscript{4}.

For DECREASE-IV, the 95\% confidence interval for the estimated number
of fabricated data points is {[}8 - 1059{]} or {[}7 - 1066{]} when based
on a point estimate or a more uncertain distribution estimate,
respectively. The minor difference between the estimates indicates that
there is a high degree of confidence that data fabrication did occur
based on the difference of the results alone. Nonetheless, the range of
potentially fabricated data points is still estimated at approximately
1000; this indicates that the summary results are insufficient to
provide more than an estimated lowerbound. This does indicate that it is
possible not all data were fabricated (i.e., \(N=1066\)), increasing the
importance of well-documented data provenance to discern between genuine
and falsified data. This affirms the conclusion of the scientific
integrity committee that the results of the DECREASE-IV trial are
``scientifically incorrect'' (p.11)\textsuperscript{3}.

The results also highlight that, despite the lack of availability of raw
data, summary results from larger samples result in more certainty about
the estimated number of fabricated data points using the inversion
method. This is due to decreased standard errors at the population level
of the effects, providing less variation in the simulations and
resulting in higher sensitivity. However, this also highlights that in
order to prevent detection it would be in the fabricators' interest to
fabricate small studies (assuming the fabricator wants to remain
undetected). This adds another factor to the unreliability of studies
with small sample sizes.
